---
title: "FP1 P2"
format: html
editor: 
  markdown: 
    wrap: 72
---

AI Usage: No AI used

Author: Sebastian Stefanowicz

Group 5: Sebastian Stefanowicz, Eric Wentz, and Natalia Morales Flores

Subject Area: movies!

Research Questions: 1. Movie adaptations: how does the relative
faithfulness of movie adaptations to their source material affect their
outcomes (critical reception, audience reception, box office, etc.)

2\. Recommendations based on user preferences and movie characteristics:
using the user preferences at that moment to recommend a movie with
similar traits, including from ratings, to genre and length in minutes,
etc.

3\. Cult Classics: classify movies with a high volume of
votes/logs/watches but low box office earnings as “cult classics”/assume
that they gained acclaim/popularity after their initial release, then
compare these films with each other and contrast them with films that
were successful in theaters.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(TMDb)

title_basics <- read_tsv("data/raw/title.basics.tsv")
```

Source: Internet Movie Database

Data description: This data frame contains basic information for IMDB's
full list of films including, title, medium (short, movie, tv show,
etc), runtime in minutes, up to 3 primary genres, R-rating, release
years, and a unique identifier.

Limitations: The primary limitation for this and all IMDB data tables is
the absence of box office and budget variables.

Variables:

tconst: unique identifier

titleType: format

primaryTitle: self explanatory

originalTitle: self explanatory

isAdult: boolean = 1 if the title is R-rated

startYear: release year

endYear: the year in which a TV show ended, NA for movies

runtimeMinutes: self explanatory

genres: a string containing up to 3 genres most closely associated with
the title separated by commas

Viz:

```{r}
title_basics_exp <- title_basics %>% filter(titleType == "movie", ) %>% mutate(decades = (as.numeric(startYear) %/% 10) * 10)

ggplot(title_basics_exp, aes(x=decades, fill=factor(isAdult))) +
  geom_bar() +
  scale_fill_manual(
    values = c("0" = "skyblue2", "1" = "orange"),
    labels = c("Not R-rated", "R-rated")
  ) +
  labs(title="Count of R-Rated and Non R-Rated Films by Decade", x="Decade", y="Count",fill=NULL) +
  theme_classic()


title_ratings <- read_tsv("data/raw/title.ratings.tsv")
```

Source: Internet Movie Database

Data description: Contains the total number of votes and average rating
for movies (identified with their unique identifiers).

Limitations: The primary limitation for this and all IMDB data tables is
the absence of box office and budget variables.

Variables:

tconst: unique identifier

averageRating: the average rating from 1-10 given to a title by users of
the IMDB site

numVotes: the total number of users who have rated a title on the IMDB
site

Viz:

```{r}
ggplot(title_ratings, aes(x=averageRating)) +
  geom_density(fill="skyblue") +
  labs(title="Density of Average IMDB User Ratings", x="Average Rating", y="Density") +
  theme_classic()
```

Join on: These, and other IMDB tables, could be easily joined on the tconst unique identifier for each movie/show.

