---
title: "LLM Classification and Preliminary Results"
format: html
---
Sebastian Stefanowicz

### Setup
```{r message=FALSE}
Sys.setenv(OPENAI_API_KEY = "insert API key here")

library(tidyverse)
library(ellmer)
library(dplyr)
library(tibble)
library(patchwork)

# movies_df_10k <- read_csv("data/raw/top_10000_movies.csv")
```

### LLM Classification using OpenAI API
```{r message=FALSE}
# Clean Data
# movies_10k_w_llm <- movies_df_10k %>% filter(release_date < "2010-01-01", budget > 10, revenue > 10)
# 
# llm_input_10k <- movies_10k_w_llm %>% select(title, release_date) %>% mutate(prompt_text = paste0(
#     "Title: ", title, ". Release date: ", release_date, "."))

# Run single-shot detection using GPT-4.1-mini with the structured output of a 1-0 binary classifier variable
# single_shot_results <- parallel_json_completion(
#   system_prompt = "Evaluate whether or not each film is widely/definitively considered to be a 'cult classic'.",
# 
#   user_prompts = llm_input_10k$prompt_text,
# 
#   type_obj = type_object(
#     llm_cult = type_string("Either 1 (a cult classic) or 0 (not a cult classic)")
#   ),
# 
#   model = "gpt-4.1-mini"
# )

# Add results to the df
# movies_10k_w_llm$llm_class_1 <- single_shot_results$llm_cult

# Repeat 5 times and aggregate results to account for anomalous results

# Create sum variable showing the number out of the 5 llm classifications that said each subject was a cult film
# Create final classification variable taking films that received a cult classification at least 4 out of 5 times
# movies_10k_w_llm <- movies_10k_w_llm %>% mutate(llm_sum = as.numeric(llm_class_1) + as.numeric(llm_class_2) + as.numeric(llm_class_3) + as.numeric(llm_class_4) + as.numeric(llm_class_5),
#                                                 llm_class_final = if_else(llm_sum >= 4, 1, 0))

# Look at distribution
# table(movies_10k_w_llm$llm_sum)

# Save results
# write.csv(movies_10k_w_llm, "data/raw/10k_pre-2010-01-01-w-llm-classification.csv")
```


### Visualized Characteristics of These Two Classification Groups
```{r message=FALSE}
movies_10k_w_llm <- read_csv("data/raw/10k_pre-2010-01-01-w-llm-classification.csv")

# Set up fill
movies_10k_w_llm <- movies_10k_w_llm |>
  mutate(
    llm_class_final = factor(llm_class_final,
                             levels = c(0, 1),
                             labels = c("No", "Yes"))
  )

# Individual plots
vote_average_plot <- ggplot(movies_10k_w_llm, aes(x = vote_average, fill = llm_class_final)) +
  geom_density(position = "identity", alpha = 0.7) +
  labs(y = NULL, x = "Average User Rating", fill = "Cult Classic Status") +
  theme_classic()

log_revenue_plot <- ggplot(movies_10k_w_llm, aes(x = log(revenue), fill = llm_class_final)) +
  geom_density(position = "identity", alpha = 0.7) +
  labs(y = NULL, x = "Log Box Office Revenue", fill = "Cult Classic Status") +
  theme_classic()

log_budget_plot <- ggplot(movies_10k_w_llm, aes(x = log(budget), fill = llm_class_final)) +
  geom_density(position = "identity", alpha = 0.7) +
  labs(y = NULL, x = "Log Production Budget", fill = "Cult Classic Status") +
  theme_classic()

runtime_plot <- ggplot(movies_10k_w_llm, aes(x = runtime, fill = llm_class_final)) +
  geom_density(position = "identity", alpha = 0.7) +
  labs(y = NULL, x = "Runtime (Minutes)", fill = "Cult Classic Status") +
  theme_classic()

# Final plot
final_plot <- (
  (vote_average_plot | runtime_plot) /
  (log_budget_plot | log_revenue_plot)
) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

final_annotated <- final_plot +
  plot_annotation(
    title = "Distribution of Movie Characteristics by LLM Cult Classification",
    caption = "Sources: TMDB and OpenAI APIs"
  )

# Save and display final plot
# ggsave("final_plot.png", final_annotated, width = 8, height = 6, dpi = 300)
final_annotated
```

Interpretation: On average cult classics, as identified by GPT-4.1-mini, have shorter runtimes, lower budgets, lower box office revenues, and more tightly concentrated average user scores than non-cult classics.


# Preliminary classification
```{r message=FALSE}
# Logistic Regression
model_logit <- glm(llm_class_final ~ vote_average + popularity + budget + revenue + runtime,
                   data = movies_10k_w_llm,
                   family = binomial)
summary(model_logit)
```

This test model corroborates our the visual differences in classification groups observed with the above plots i.e. cult and non-cult films appear to on average differ in user score, budget, revenue, and runtime.

